---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```


```{r}
#| echo: false
#| include: false 

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               marginaleffects,
               texreg, 
               ggdist,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)


options(scipen=999)
rm(list = ls())

load(file = here("input/data/proc/df_study1_long_t7.RData"))

# Generate analytical sample

df_study1 <- df_study1_long_t7 %>%
  select(idencuesta,
         ola,
         ponderador_long_total,
         segmento,
         estrato,
         just_health,
         just_pension,
         just_educ,
         mjp, 
         perc_inequality, 
         merit_effort, 
         merit_talent, 
         educ,
         quintil1,
         sex,
         age,
         ess, 
         ideo) %>% 
  na.omit() %>% 
  mutate(ola = case_when(ola == 1 ~ 1,
                         ola == 2 ~ 2, 
                         ola == 3 ~ 3,
                         ola == 4 ~ 4,
                         ola == 6 ~ 5,
                         ola == 7 ~ 6)) %>% 
  mutate(ola = as.factor(ola),
         ola_num = as.numeric(ola),
         ola_2=as.numeric(ola)^2)

df_study1 <- df_study1 %>%
  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
  ungroup()

df_study1 <- df_study1 %>% filter(n_participaciones>1)

# Corregir etiquetas

df_study1$just_health <- sjlabelled::set_label(df_study1$just_health, 
                        label = "Health distributive justice")

df_study1$just_pension <- sjlabelled::set_label(df_study1$just_pension, 
                        label = "Pension distributive justice")

df_study1$just_educ <- sjlabelled::set_label(df_study1$just_educ, 
                        label = "Education distributive justice")

df_study1$merit_effort <- sjlabelled::set_label(df_study1$merit_effort, 
                        label = "People are rewarded for their efforts")

df_study1$merit_talent <- sjlabelled::set_label(df_study1$merit_talent, 
                        label = "People are rewarded for their intelligence")

```

# Data, Variables and Methods

## Data

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a nationally representative panel study of the urban adult population in Chile, conducted annually between 2016 and 2023. Designed to examine individuals’ attitudes, emotions, and behaviors regarding social conflict and cohesion, ELSOC employs a probabilistic, stratified, clustered, and multistage sampling design covering both major urban centers and smaller cities. The sampling frame was proportionally stratified into six categories of urban population size (e.g., large and small cities), followed by the random selection of households within 1,067 city blocks. The target population includes men and women aged 18 to 75 who are habitual residents of private dwellings.

The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic. This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was excluded because a reduced version of the questionnaire omitted key variables of interest. Between waves 1 and 6, panel attrition reached 40%, resulting in a final two-level sample comprising N = 8,643 observations nested within N = 1,687 individuals. Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response. Further details on sampling, attrition, and weighting procedures are available at https://coes.cl/encuesta-panel/, and the dataset is publicly accessible at https://dataverse.harvard.edu/dataverse/elsoc.

## Variables

**Market justice preferences**: the dependent variable in this study is preferences for market justice. This construct is operationalized through three items that capture how strongly individuals justify conditioning access to core services---healthcare, pensions, and education--- basen on individual income. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84). This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).

```{r}
#| label: tbl-summary1
#| tbl-cap: "Dependent variables for the first wave (2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false


t1 <- df_study1 %>% 
  filter(ola == 1) %>% 
  select(just_health, just_pension, just_educ, mjp) 

df<-dfSummary(t1,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)
```



**Perception of economic inequality**: this variable is measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999]. This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy. Specifically, it relies on the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011]. Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum. This measure includes a logarithmic term in order to adjust income magnitudes (usually fewer cases with high income):

$$
\text{perceived wage gap} = \log_{10}\left(\frac{\text{perceived salary of a large-company president}}{\text{perceived salary of an unskilled worker}}\right)
$$

**Perception of Meritocracy**: this variable is operationalized through two components, namely effort and talent [@young_rise_1962]. The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills". In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").


Table 2 shows the independent variables used, their response categories and their frequencies.

```{r}
#| label: tbl-summary2
#| tbl-cap: "Independent variables ELSOC survey (descriptives for first wave 2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false


t2 <- df_study1 %>% 
  filter(ola == 1) %>% 
  select(perc_inequality, merit_effort, merit_talent) 

df<-dfSummary(t2,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)
```

*Controls*

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in ranges), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024]. Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (measured through a scale from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015]. Descriptive statistics for the control variables can be found in the supplementary material.

## Methods

Given the data's hierarchical structure, in which observations are nested in survey waves, we employ longitudinal multilevel linear models [@singer_applied_2009]. In a panel-data framework, within-person effects capture how shifts in individual-level variables across waves are associated with variations in market justice preferences. By contrast, between-person effects focus on differences among individuals, explaining how long-term (or average) values relate to overall levels of market justice preferences.

To estimate within-person effects, we use group-mean centering, where each respondent functions as the "group" (i.e., observations nested within persons). Meanwhile, the between-person effects are derived from each individual's average on these variables, calculated across the waves of panel data.

All the analyses were conducted using R software and the *lme4* package [@bates_fitting_2015].
